\textbf{Microbiota predict \emph{Clostridium difficile} severity in germ-free mice colonized with human feces}

### 2021-12-14 submission/manuscript.tex
update manuscript with pds edits round 2

\hypertarget{the-gut-microbiota-potentiates-clostridiodes-difficile-infection-severity.}{%
\section{\texorpdfstring{The gut microbiota potentiates
\emph{Clostridiodes difficile} infection
severity.}{The gut microbiota potentiates Clostridiodes difficile infection severity.}}\label{the-gut-microbiota-potentiates-clostridiodes-difficile-infection-severity.}}

@@

Nicholas A. Lesniak\(^1\), Alyxandria M. Schubert\(^1\), Kaitlyn
Flynn\(^1\), Hamide Sinani\(^1\), Kathyrn A. Eaton\(^1\), Ingrid L.
Bergin\(^2\), Patrick D. Schloss\(^{1,\dagger}\)

@@

\(\dagger\) To whom correspondence should be addressed:
\href{mailto:pschloss@umich.edu}{\nolinkurl{pschloss@umich.edu}}\\
1. Department of Microbiology and Immunology, University of Michigan,
Ann Arbor, MI\\
2. Unit for Laboratory Animal Medicine, University of Michigan, Ann
Arbor, MI

@@

\emph{Clostridioides difficile} infection has become more common and
severe in the last few decades. Patient age, white blood cell count,
creatinine levels as well as \emph{C. difficile} ribotype and the
presence of toxin genes have been associated with disease severity.
However, models based on those data to identify patients with CDI that
will develop severe disease have not been robust enough to be broadly
applied. These models are input from our understanding of \emph{C.
difficile} interactions with the gut environment. The gut bacteria are a
key determinant in \emph{C. difficile} infection. CDI is dependent on
perturbations in the bacterial community in order to colonize the gut.
The gut microbiota also impairs \emph{C. difficile} colonization through
bile acid metabolism, nutrient consumption and bacteriocin production.
However, it is unclear if the gut bacteria affect the disease severity
resulting from that colonization. Here we have shown gut bacterial
communities contribute to disease severity variation. We created diverse
gut communities by colonizing germ-free mice with human fecal
communities. They were then infected with \emph{C. difficile} which
resulted in differed disease severities. The severity grouped by the
human fecal community they received. Generally, facultative anaerobes
with pathogenic potential, such as Escherichia, Helicobacter and
Klebsiella, were associated with more severe disease. Bacterial groups
associated with dietary fiber degradation, such as Coprobacillus, were
associated with reduced disease severity. Lastly, we showed communities
that resulted in either low or high histopathologic scores or severe
disease had the same outcome when infected with a different \emph{C.
difficile} isolate.

@@

Clostridioides difficile infection (CDI) causes a range of disease from
asymptomatic to mild diarrhea to severe outcomes of recurrent infections
and even death. Models that predict severity and treatment decisions are
based on clinical factors and \emph{C. difficile} capabilities. But
currently the gut microbiome, the primary protector against CDI, has not
been accounted for its effect on CDI disease severity. We demonstrated
variation in the microbiome of mice colonized with human feces was
sufficient to result in a range of disease severities. Our results show
groups of bacteria contribute to the development of disease in \emph{C.
difficile} infections. Gut bacterial community data from patients with
CDI could improve our ability to identify patients more at risk for
severe disease and improve interventions which treat the gut bacteria to
reduce host damage.

@@

\emph{Clostridioides difficile} infections (CDI) have been increasing in
incidence and severity since \emph{C. difficile} was first identified as
a pathogen. It was first identified in 1931 in a healthy fecal community
{[}CITATION{]}. But then in the late 1970s, following the introduction
of broad-spectrum antibiotics, \emph{C. difficile} was identified as the
cause of antibiotic-associated psuedomembranous colitis {[}CITATION{]}.
Beginning in the early 2000s, CDI became more problematic with
increasing incidences and severity {[}CITATION{]}. Disease severities
can range from asymptomatic to mild diarrhea to toxic megacolon and
death. This severity has been associated with characteristics of
\emph{C. difficile} ribotypes and host factors. Over time more
pathogenic ribotypes have emerged, such as ribotype 027 in the early
2000s. The presence of the genes for Toxin A/B or fluoroquinolon
resistence has been associated with more severe disease {[}CITATION{]}.
Binary toxin, while alone insufficient to cause damage on it own, when
in the presence of Toxin A/B has been associated with increased
virulence {[}CITATION{]}. Mutations in tcd C, the negative regulator of
of toxin transcription, also increased the virulence of \emph{C.
difficile} {[}CITATION{]}. In addition to the pathogenic potential of
the infecting strain, the response by the host also affects the disease
severity. Delayed or reduced production of antibodies IgG and IgA can
result in more severe disease {[}CITATION{]}. Increased neutrophil
inflitration can lead to worse outcomes {[}CITATION{]}. In addition to
the immune response, increased density of the toxin receptor on the host
cells can increase the severity of the infection {[}CITATION{]}. While
there is a thorough understanding of many of the mechanisms driving in
CDI incidence and disease severity, there still is not effective tools
to reduce the risk of severe disease.

Models have been developed to identify patients at risk for severe CDI
but have not been robust for broad application (1). These models utilize
patient related data, such as patient age, white blood cell count, serum
albumin levels, and creatinine levels, to predict disease severity. In
their initial publication, these models perform well enough for
publication. However when Perry et al applied many of the most current
models to a multi-center external validation, their performance suffered
and often had more false positives than true positives (3). Thus, there
seems to be a missing factor contributing to disease severity.

Missing from these models are the gut bacteria which also interact with
\emph{C. difficile} during the course of the infection. The indigenous
bacteria of a healthy intestinal community provide a protective barrier
preventing \emph{C. difficile} from infecting the gut. Only after this
community is disrupted from a range of pertubations, such as many
classes of antibiotics, medications, or even dietary changes, can
\emph{C. difficile} infect the intestine. Once established, the gut
bacteria can either promote or inhibit \emph{C. difficile} through
producing molecules or modifying the environment {[}Abbas2020{]}. Bile
acids metabolized by the gut bacteria can inhibit \emph{C. difficile}
growth and affect toxin production (5). Bacteria in the gut also can
compete more directly with \emph{C. difficile} through antibiotic
production or nutrient consumption (7). To eliminate CDI, the primary
treatment is a course of antibiotics. Patients who become reinfected and
recalitrant to antibiotics, a fecal microbiota transplantion often is
sufficient to restore the gut community to eliminate \emph{C.
difficile}. While there has been many studies to investigate the
interaction between the gut bacteria and \emph{C. difficile} during the
infection, it has not been demostrated the bacterial community can
modulate the disease severity of CDI. Recently, Tomkovich et al showed
the variation in the bacterial communtiies between mice from different
colonies and vendors was sufficient to cause differences in the temporal
dynamics of the \emph{C. difficile} infection (9). Since the gut
bacteria interact with \emph{C. difficile} throughout the infection and
community difference resulted in different colonization dynamics, we
hypothesized that the gut bacteria contribute to the variation in CDI
disease severity. Here, we colonized germ-free C57BL/6 mice with human
fecal communities to test if the variablity in the gut bacteria can
produce and explain variation in the disease severity caused by a single
\emph{C. difficile} isolate.

@@

\textbf{Germ-free mice colonized with human fecal commmunities have
diverse and unique community structures.} Based on our previous
observation mice with variation in their microbiota showed differences
in colonization of \emph{C. difficile} (9), we sought to explore the
effect communties with greater variation had on \emph{C. difficile}
challenge. To produce communities with greater variation than mouse
colonies with their indigenous communties, we used human fecal
communities to establish a community in the intestine of germ-free mice.
We inoculated a cage of germ-free C57BL/6 mice (2-4 mice per cage) with
homogenized feces from one of 15 different human donors via oral gavage.
The fecal community was allowed to colonize and stabilize for two weeks
post inoculation {[}CITATION?{]}. We then assessed the gut bacterial
communities by sequencing the V4 region of the 16S rRNA gene extracted
from fecal pellets (Figure 1A). The communities established in the mice
group most closely to other mice recieving the same human fecal donor
community than to either their respective donor or any mice who received
stool from a different donor (Figure 1B). Most of the communities were
primairily composed of populations of \emph{Clostridia},
\emph{Bacteroidia}, \emph{Erysipelotrichia}, \emph{Bacilli}, and
\emph{Gammaproteobacteria}. However, each group of mice harbored unique
combination of class abunadances in their gut bacterial communities.
\emph{Any alpha? compare to murine variation? Number of unique otus?}

\textbf{\emph{C. difficile} is able to colonize mice without
perturbation.} Our goal was to test the effect of the variation in the
microbiota on \emph{C. difficile} infection. A typical mouse model of
CDI requires pretreatment with antibiotics such as clindamycin to become
susceptible to \emph{C. difficile} colonization {[}CITATION{]}. However,
we wanted to avoid modifying the communties with an antibiotic to
maintain their community structures and differences. Since these
communties came from human donors which may have become susceptible
through their own exposures, we decided to test if \emph{C. difficile}
was able to colonize these communities without any perturbation. After
two weeks post incoluation with human fecal communities, the mice were
then challenged with 10\^{}\textbf{4?} \emph{C. difficile} ribotype 027
isolate 431 spores (\textbf{Add schematic?}) (10). The mice were
followed for 10 days post challenge and their stool was collected and
plated for \emph{C. difficile} CFU to detect colonization level. We
hypothesized that \emph{C. difficile} would most likely be able to
colonize the mice with communities that came from donors which had a
history of antibiotic use, one of the primary risk factors for
developing CDI (TABLE, CITATION). Suprisingly, communities from all
donors were colonized (Figure 2). The only two mice that were able to
resist \emph{C. difficile} colonization both recieved their community
from the same donor. Therefore, the transplanted human fecal communties
were susceptible to \emph{C. difficile} colonization without the need
for any perturbation.

\textbf{Infection seveity varies by initial community structure.} After
challenging the mice with \emph{C. difficile}, we looked at the effect
of the community variation on the outcome from the \emph{C. difficile}
infection. We followed the mice for signs of disease as well as \emph{C.
difficile} colonization and toxin production for 10 days post challenge.
Some mice were not initally colonized on the day after \emph{C.
difficile} challenge but became colonized with a few days post challenge
(\textbf{SUPP FIG?}). A subset of mice, all that recieved their comunity
from one of 6 donors, suffered from severe disease, which became highly
colonized one day post challenge and moribund within 3 days post
challenge. The remaining mice, except for one cage, maintained \emph{C.
difficile} colonization through the end of the experiment (Figure 2).
After the mice were humanely euthanised at the endpoint, day 10 post
challenge or earlier if they became moribund, their cecal tissue was
collected and scored for histological signs of disease and fecal samples
were assayed for toxin (Figure 3). Overall, there was greater toxin
activity detected in the stool of the moribund mice (\emph{P} = 0.003).
However, when looking at each group of mice, we saw there was a range in
toxin activity for both the mice with non-severe disease and severe
disease (Figure 3A). Some donor groups that maintained colonization
without severe disease had similarlly high activity levels as was
detected in the mice with severe disease. Additionally, some mice with
severe disease did not have toxin activity detected in their stool.
These oberserved variations in toxin activity and relation to disease
severity have been previously characterized (11). Next we examined the
cecal tissue for histopathological damage. For mice with severe disease,
we observed high levels of epithelial damage, tissue edema and
inflammation (Figure SXXXX), similar to previously reported
histopathologic findings for \emph{C. difficile} ribotype 027 (13). As
observed with toxin activity, the mice with severe disease had higher
histopathologic scores than the mice with non-severe disease (\emph{P} =
3.0e-9). However, unlike the toxin activity, all moribund mice had
consistently high histopathologic scores. The mice which maintained
persistent population size, we saw a range in tissue damage from no
signs of disease to nearly the same level as the mice with severe
disease, which grouped by community donor. Together, the toxin activity,
histopathologic score, and moribundity showed variation across the
donors but were largely consistent within each group of mice recieving
the same human donor fecal community.

\textbf{Populations of the microbial communities explain variation in
CDI disease severity.} Since the disease charateristics grouped by human
donor fecal community, we next investigated the community composition
for its ability to explain the variation in disease severity. First, we
tested for associations between taxonomic groups and the outcomes of the
\emph{C. difficile} infection. We used the linear discriminant analysis
(LDA) effect size (LEfSe) method to identify differences in the genera
of the initial bacterial communities between the severity level and
toxin production. We split the mice into groups by severity level based
on their if they became moribund (severe), or if their clinical score
was above 5 (moderate) or below 5 (mild). This analysis revealed 20
genera that were significantly different by the disease severity (Figure
4A). The presence of populations of \emph{Turicibacter},
\emph{Streptococcus}, \emph{Staphylococcus}, \emph{Pseudomonas},
\emph{Phocaeicola}, \emph{Parabacteroides}, \emph{Bacteroides}, and
\emph{Escherichia/Shigella} were detected at higher levels in the mice
that developed severe disease. Populations of \emph{Anaerotignum},
\emph{Coprobacillus}, \emph{Enterocloster}, and \emph{Murimonas} were
increased in the mice that would experience mild disease. There were
many genera that did not have a correlation of relative abundance and
disease severity, such as the populations of \emph{Monoglobus} and
\emph{Hungatella}, which were decreased or increased, respectively, in
only the group of mice that would develop moderate disease. To
understand these cases better we used LEfSe to identify the genera
differentiating the communities that we detected toxin in from those
that we did not (Figure 4B). There were many genera the were associated
with toxin production as we would expect from their association with
severe disease, such as populations of \emph{Escherichia/Shigella} and
\emph{Bacteroides}. Likewise, there were genera such as
\emph{Anaerotignum}, \emph{Enterocloster}, and \emph{Murimonas} that
were higher in communities that had mild disease and no toxin was
produced. However, communities without a linear trend between relative
abundance and disease severity can be better understood by their
association with toxin production. Populations of \emph{Hungatella} were
increased in the group with moderate disease but not with the production
of toxin. Lastly, we looked for associations between the histopathologic
score and the populations of genera at that same time (Figure 4C). The
populatons of \emph{Bacteroides} at the end matched the trend observed
in the initial community, increased population correlated with increased
histopathologic score. Populations of \emph{Klebsiella} and
\emph{Prevotellaceae} positively correlated with clinic score and were
increased in the group that produced toxin. These tests have identified
which individual populations of bacteria that associated with each
disease severity.

We next wanted to understand how the populations in the context of each
other lead to the different disease severities. We trained random forest
models with the intital bacterial community relative abundance data at
each taxonomic rank to predict toxin production, severe disease or final
histopathologic score. Overall for predicting toxin production,
microbial populations aggregated by their phylum level classificiation
performed as well as models using lower taxonomic ranks (AUROC = 0.83,
Figure S2). In the initial community populations of
\emph{Verrucomicrobia}, \emph{Campilobacterota} and
\emph{Proteobacteria} contributed the most to the correct prediction of
toxin production (Figure 5A). \emph{C. difficile} was more likely to
produce toxin when the initial community had smaller populations of
\emph{Verrucomicrobia} and \emph{Campilobacterota} and had larger
popluations of \emph{Proteobacteria}. Next, we assessed the ability of
the community to predict severe disease. Class rank classification was
sufficient to predict if the mouse would succumb to the infection before
the end of the experiment (AUROC = 0.91, Figure S2). Groups of bacteria
belonging to \emph{Bacilli}, \emph{Firmicutes}, and
\emph{Erysipelotrichia} contributed the most to the correct prediction.
Larger populations of \emph{Bacilli} and \emph{Firmicutes} and reduced
populations of \emph{Erysipelotrichia} were more likely to result in
moribundity (Figure 5B). Aside from \emph{Erysipelotrichia}, the only
other population decreased in moribund mice was \emph{Clostridia}, all
other features had increased populations in the moribund mice. Lastly,
the initial community was able to predict if the mice would have a
histopathologic score above or below the median score (median
histopathologic score = 5) with the genera relative abundances (AUROC =
0.99, Figure S2). There were no genera with a significantly greater
effect on the model performance than any others, indicating the model is
reliant on many genera for the correct prediction. The model used some
of the genera identified in the LEfSe analysis, such as populations of
\emph{Coprobacillus}, \emph{Anaerostipes}, and \emph{Hungatella}.
Communities with greater populations of \emph{Hungatella},
\emph{Eggerthella}, \emph{Bifidobacterium}, \emph{Duncaniella} and
\emph{Neisseria} were more likely to have greater histopathologic
scores.

\textbf{Disease severity consistent by donor community across
strains/isolates.} We used a single \emph{C. difficile} isolate,
ribotype 027 isolate 431 to characterize the range of disease severity
and the features of the bacterial community associated with severity
(10). Since we had used the same mice and \emph{C. difficile} isolate in
our first set of experiemnts, we next wanted to test if the effect of
the community would apply to other \emph{C. difficile} isolates. We
selected three communities based on the result from our analysis thus
far to select one community we would expect to result in a low clinical
score (\textless{} 5), a high clinical score (\textgreater{} 5), and one
with a high clinical score which becomes moribund. Using genus level
data to select the communities, we selected communities based on the
relative abundance patterns of \emph{Sellimonas} and \emph{Anaerotignum}
(higher abundance in mice with a low clinical score),
\emph{Lachnospiracea} (lower abundance in moribund mice),
\emph{Hungatella} and \emph{Eisenbergiella} (higher abundance in
moderate mice and lower in moribund mice), Clostridium sensu stricto
(higher in moribund mice by both LEfSe analysis and RF model) and
\emph{Enterocloster} (lower in moribund mice and a negative linear trend
with disease severity by LEfSe analysis) (Figure S3). With this set of
human fecal communities selected, we inoculated mice via oral gavage of
the human fecal slurry and allowed two weeks for the a community to
establish. We then challenged the mice with a different \emph{C.
difficile} (10). The disease severity for this set of experiments
closely matched that in our first set of experiments with \emph{C.
difficile} isolate 431 (Figure 6).

@@

Challenging mice colonized with different human fecal communities with a
single \emph{C. difficile} isolate allowed us to identify the effect of
microbiome variation on \emph{C. difficile} infection disease severity.
Our LEfSe analyses and random forest models revealed the relationships
between the bacterial community members and the degree of disease
severity. We used the relative abundance patterns of community members
with strong associations with disease severity to select communities to
test if the observed disease severity is conserved when other \emph{C.
difficile} isolates are used. The disease outcome from the \emph{C.
difficile} chalenge with the new isolate matched the expected disease
severities. Overall, these results have demonstrated that the bacterial
community is capable of modulating the disease severity of \emph{C.
difficile} infection.

>>>>

\hypertarget{the-gut-bacterial-community-potentiates-clostridioides-difficile-infection-severity.}{%
\section{\texorpdfstring{The gut bacterial community potentiates
\emph{Clostridioides difficile} infection
severity.}{The gut bacterial community potentiates Clostridioides difficile infection severity.}}\label{the-gut-bacterial-community-potentiates-clostridioides-difficile-infection-severity.}}

@@

Nicholas A. Lesniak\(^1\), Alyxandria M. Schubert\(^1\), Kaitlyn
Flynn\(^1\), Jhansi L Leslie\(^1\), Hamide Sinani\(^1\), Ingrid L.
Bergin\(^3\), Vincent B Young\(^2\), Patrick D. Schloss\(^{1,\dagger}\)

@@

\(\dagger\) To whom correspondence should be addressed:
\href{mailto:pschloss@umich.edu}{\nolinkurl{pschloss@umich.edu}}\\
1. Department of Microbiology and Immunology, University of Michigan,
Ann Arbor, MI\\
2. Division of Infectious Diseases, Department of Internal Medicine,
University of Michigan Medical School, Ann Arbor, MI\\
3. Unit for Laboratory Animal Medicine, University of Michigan, Ann
Arbor, MI

@@

The severity of \emph{Clostridioides difficile} infections (CDI) has
increased over the last few decades. Patient age, white blood cell
count, creatinine levels as well as \emph{C. difficile} ribotype and
toxin genes have been associated with disease severity. However, it is
unclear whether there is an associate between the gut microbiota disease
severity. The gut microbiota is known to interact with \emph{C.
difficile} during infection. Perturbations to the gut microbiota are
necessary for \emph{C. difficile} to colonize the gut. The gut
microbiota is thought to impair \emph{C. difficile} colonization through
bile acid metabolism, nutrient consumption and bacteriocin production.
It has not been shown whether the gut microbiota can affect CDI disease
severity. Here we demonstrate that gut bacterial communities can
contribute to disease severity. We derived diverse gut communities by
colonizing germ-free mice with different human fecal communities. The
mice were then infected with a single \emph{C. difficile} ribotype 027
clinical isolate which resulted in moribundity and histopathologic
differences. The variation in severity was associated with the human
fecal community that the mice received. Generally, facultative anaerobes
with pathogenic potential, such as \emph{Escherichia},
\emph{Helicobacter}, and \emph{Klebsiella}, were associated with more
severe outcomes. Bacterial groups associated with fiber degradation,
such as \emph{Coprobacillus}, were associated with less severe outcomes.
Finally, we showed that, when infected with a different \emph{C.
difficile} isolate, the outcome of the infection was similar between the
mice that received the same fecal donor community. These data indicate
that gut bacteria can influence CDI disease severity.

@@

\emph{Clostridioides difficile} infections (CDI) can be asymptomatic or
range in severity from mild diarrhea to severe outcomes including
surgical interventions and death. Models that predict severity and guide
treatment decisions are based on clinical factors and \emph{C.
difficile} characteristics. Although the gut microbiome plays a role in
protecting against CDI, it has not been investigated for its effect on
CDI disease severity or incorporated into attempts to predict disease
severity. We demonstrated that variation in the microbiome of mice
colonized with human feces yielded a range of disease outcomes. These
results revealed groups of bacteria associated with both severe and mild
\emph{C. difficile} infection outcomes. Gut bacterial community data
from patients with CDI could improve our ability to identify patients at
risk of developing more severe disease and improve interventions which
treat the gut bacteria to reduce host damage.

@@

\emph{Clostridioides difficile} infections (CDI) have increased in
incidence and severity since \emph{C. difficile} was first identified as
the cause of antibiotic-associated pseudomembranous colitis (1). CDI
disease severity can range from asymptomatic to mild diarrhea to toxic
megacolon and death. The Infectious Diseases Society of America (IDSA)
and Society for Healthcare Epidemiology of America (SHEA) guidelines
categorize severe disease based on white blood cell count greater than
15,000 cells/mm\(^3\) and/or a serum creatinine greater than 1.5 mg/dL,
and complicated disease based on systolic blood pressure below 90,
ileus, or toxic megacolon (2). Since these measures are only present
once the disease has advanced, the IDSA classification does not allow
for identification of patients at risk of severe CDI. Prediction models
have been developed to score a patient's risk for severe CDI outcomes
based on clinical factors but have not been robust for broad application
(3--6). Thus, we have limited ability to prevent patients from
developing severe CDI. Ideally, we could identify patients that are
high-risk of severe disease so their treatments could be adjusted to
halt the infection before the disease worsens.

Missing from CDI severity prediction models are the effects of the
indigenous gut bacteria. \emph{C. difficile} interacts with the gut
community in many ways. The indigenous bacteria of a healthy intestinal
community provide a protective barrier preventing \emph{C. difficile}
from infecting the gut. A range of mechanisms can disrupt this barrier,
including antibiotics, medications, or dietary changes, and lead to
increased susceptibility to CDI (7--9). Once in the gut, the indigenous
bacteria can either promote or inhibit \emph{C. difficile} through
producing molecules or modifying the environment (10, 11). Bile acids
metabolized by the gut bacteria can inhibit \emph{C. difficile} growth
and affect toxin production (12, 13). Bacteria in the gut also can
compete more directly with \emph{C. difficile} through antibiotic
production or nutrient consumption (14, 15). While the relationship
between the gut bacteria and \emph{C. difficile} has been established,
the effect the gut bacteria can have on CDI disease severity is unclear.

Studies investigating the severity of CDI disease involving the
microbiome have not shown clear evidence of a relationship between the
microbiome and disease severity. These studies were not setup to
identify a relationship between the microbiome and disease severity.
Studies that have used clinical data cannot control for host, microbiome
or \emph{C. difficile} ribotype variation (16). While disease severity
varies, it cannot be attributed to specifically the microbiome since all
factors vary. Murine experiments typically use a single mouse colony and
different \emph{C. difficile} ribotypes to create severity differences
(3). Recently, our group has begun uncovering the effect microbiome
variation has on \emph{C. difficile} infection. We showed the variation
in the bacterial communities between mice from different mouse colonies
resulted in different clearance rates of \emph{C. difficile} (17). We
also showed varied ability of mice to spontaneously eliminate \emph{C.
difficile} infection when they were treated with different antibiotics
prior to \emph{C. difficile} challenge (18). However, there have not
been any studies investigating whether variation in the microbiome
affects the resultant disease severity of \emph{C. difficile}
infections.

Here, we sought to determine whether the gut bacteria contribute to
disease severity. We hypothesized that since the gut bacteria affect the
metabolism of \emph{C. difficile} and its infection dynamics, it can
also affect the disease severity of the infection. To test this
hypothesis, we colonized germ-free C57BL/6 mice with human fecal
communities to create varied gut communities. We then challenged the
mice with a single \emph{C. difficile} isolate, ribotype 027 (RT027)
human isolate 431, and followed the mice for the development of severe
outcomes of moribundity and histopathologic cecal tissue damage. Since
the murine host and \emph{C. difficile} isolate were the same and only
the gut community varied, the variation in disease severity we observed
was attributable to the microbiome.

@@

\textbf{\emph{C. difficile} is able to infect germ-free mice colonized
with human fecal microbial communities without perturbation.} To produce
gut microbiomes with greater variation than those found in conventional
mouse colonies, we colonized germ-free mice with bacteria from human
feces (19). We inoculated germ-free C57BL/6 mice with homogenized feces
from one of 15 human fecal samples via oral gavage. The human fecal
samples were selected because they represented diverse community
structures based on community clustering (20). The gut communities were
allowed to establish themselves for two weeks post inoculation (21). We
then surveyed the bacterial members of the gut communities by 16S rRNA
gene sequencing of fecal pellets (Figure 1A). The bacterial communities
from each mouse grouped more closely to those communities from mice that
received the same human fecal donor community than to the mice who
received a different human fecal donor community (Figure 1B). The
communities were primarily composed of populations of \emph{Clostridia},
\emph{Bacteroidia}, \emph{Erysipelotrichia}, \emph{Bacilli}, and
\emph{Gammaproteobacteria}. However, the gut bacterial communities of
each donor group of mice harbored unique relative abundance
distributions of the shared bacterial classes.

Next, we tested this set of mice with their human-derived gut microbial
communities for susceptibility to \emph{C. difficile} infection. A
typical mouse model of CDI requires pre-treatment with antibiotics, such
as clindamycin, to become susceptible to \emph{C. difficile}
colonization (22, 23). However, we wanted to avoid modifying the
communities with an antibiotic to maintain their unique microbial
compositions and ecological relationships. Since these communities came
from people who had increased risk of CDI, such as hospitalization or
recent antibiotic use (20), we decided to test if \emph{C. difficile}
was able to infect these mice without any antibiotic perturbation. We
hypothesized that \emph{C. difficile} would be able to colonize a subset
of mice whose communities were susceptible from perturbations either the
donor experienced or from the ecological selection when colonizing the
murine gut. Mice were challenged with 10\(^{3}\) \emph{C. difficile}
RT027 human isolate 431 spores. The mice were followed for 10 days
post-challenge and their stool was collected and plated for \emph{C.
difficile} colony forming units (CFU) to determine the extent of the
infection. Surprisingly, mice from all donors were colonized (Figure 2).
Only two mice were able to resist \emph{C. difficile} infection and both
received their community from the same donor, N3. By colonizing
germ-free mice with different human fecal communities, we were able to
generate diverse gut communities in mice, which were susceptible to
\emph{C. difficile} infection without further modification of the gut
community.

\textbf{Infection severity varies by initial community.} After we
challenged the mice with \emph{C. difficile}, we investigated the
outcome from the infection and its relationship to the initial
community. We followed the mice for 10 days post-challenge for
colonization density, toxin production, and mortality. Seven mice, from
Donors N1, N2 and N5, were not colonized on the day after \emph{C.
difficile} challenge but were infected (\textgreater10\(^{6}\)) by the
end of the experiment. All mice that received their community from Donor
M1 through M6 succumbed to the infection and became moribund within 3
days post-challenge. The remaining mice, except the uninfected Donor N3
mice, maintained \emph{C. difficile} infection through the end of the
experiment (Figure 2). At 10 days post-challenge, or earlier for the
moribund mice, mice were euthanised and fecal pellets were assayed for
toxin activity and cecal tissue was collected and scored for
histopathologic signs of disease (Figure 3). Overall, there was greater
toxin activity detected in the stool of the moribund mice (\emph{P} =
0.003). However, when looking at each group of mice, we observed a range
in toxin activity for both the moribund and non-moribund mice (Figure
3A). Non-moribund mice from Donors N4 through N9 had comparable toxin
activity as the moribund mice. Additionally, not all moribund mice had
toxin activity detected in their stool. This discordance between toxin
activity and disease severity has been previously described (24, 25).
Next we examined the cecal tissue for histopathologic damage. Moribund
mice had high levels of epithelial damage, tissue edema, and
inflammation (Figure S1) similar to previously reported histopathologic
findings for \emph{C. difficile} RT027 (26). As observed with toxin
activity, the moribund mice had higher histopathologic scores than the
non-moribund mice (\emph{P} \textless{} 0.001). However, unlike the
toxin activity, all moribund mice had consistently high histopathologic
summary scores (Figure 3B). The non-moribund mice, Donor groups N1
through N9, had a range in tissue damage from none detected to similar
levels as the moribund mice, which grouped by community donor. Together,
the toxin activity, histopathologic score, and moribundity showed
variation across the donor groups but were largely consistent within
each donor group.

\textbf{Microbial community members explain variation in CDI severity.}
We next interrogated the bacterial communities at the time of \emph{C.
difficile} challenge (day 0) for their relationship to infection
outcomes using linear discriminant analysis (LDA) effect size (LEfSe)
analysis to identify individual bacterial populations that could explain
the variation in disease severity. We split the mice into groups by
severity level based on their moribundity and histopathologic score. We
dichotomized the histopathologic scores into high and low groups by
splitting on the median score. This analysis revealed 20 genera that
were significantly different by the disease severity (Figure 4A).
Bacterial genera \emph{Turicibacter}, \emph{Streptococcus},
\emph{Staphylococcus}, \emph{Pseudomonas}, \emph{Phocaeicola},
\emph{Parabacteroides}, \emph{Bacteroides}, and
\emph{Escherichia/Shigella} were detected at higher relative abundances
in the mice that became moribund. Populations of \emph{Anaerotignum},
\emph{Coprobacillus}, \emph{Enterocloster}, and \emph{Murimonas} were
more abundant in the non-moribund mice that would develop only low
intestinal injury. To understand the role of toxin activity in disease
severity, we applied LEfSe to identify the genera most likely to explain
the observed differences in toxin activity (Figure 4B). Many genera that
associated with toxin production were also associated with moribundity,
such as populations of \emph{Escherichia/Shigella} and
\emph{Bacteroides}. Likewise, there were genera such as
\emph{Anaerotignum}, \emph{Enterocloster}, and \emph{Murimonas} that
were associated with no toxin production that also exhibited greater
relative abundance in communities from non-moribund mice with a low
histopathologic score. Lastly, we tested for correlations between the
endpoint relative abundances of bacterial operational taxonomic units
(OTUs) and the histopathologic summary score (Figure 4C). The endpoint
relative abundance of \emph{Bacteroides} was positively correlated with
histopathologic score, as its day 0 relative abundance did with disease
severity (Figure 4A). Populations of \emph{Klebsiella} and
\emph{Prevotellaceae} were positively correlated with the
histopathologic score and were increased in the group of mice with
detectable toxin. This analysis identified bacterial genera that were
associated with the variation in moribundity, histopathologic score, and
toxin activity of CDI.

We next determined whether collectively, bacterial community membership
and relative abundance could be predictive of the CDI disease outcome.
We trained random forest models with bacterial community relative
abundance data from the day of colonization at each taxonomic rank to
predict toxin production, moribundity, and day 10 post-challenge
histopathologic summary score. For predicting if toxin would be
produced, microbial populations aggregated by phylum rank classification
performed similarly as models using lower taxonomic ranks (AUROC = 0.83,
Figure S2). \emph{C. difficile} was more likely to produce toxin when
the community it infected had less abundant populations of
\emph{Verrucomicrobia} and \emph{Campilobacterota} and had more abundant
populations of \emph{Proteobacteria} (Figure 5A). Next, we assessed the
ability of the community to predict moribundity. Bacteria grouped by
class rank classification was sufficient to predict which mice would
succumb to the infection before the end of the experiment (AUROC = 0.91,
Figure S2). The features with the greatest effect showed that
communities with greater populations of bacteria belonging to
\emph{Bacilli} and \emph{Firmicutes} and reduced populations of
\emph{Erysipelotrichia} were more likely to result in moribundity
(Figure 5B). Only one other class of bacteria was decreased in moribund
mice, a group of unclassified \emph{Clostridia}. Lastly, the relative
abundances of genera were able to predict a high or low histopathologic
score (histopathologic scores were dichotomized as in previous analysis,
AUROC = 0.99, Figure S2). No genera had a significantly greater effect
on the model performance than any others, indicating the model was
reliant on many genera for the correct prediction. The model used some
of the genera identified in the LEfSe analysis, such as
\emph{Coprobacillus}, \emph{Anaerostipes}, and \emph{Hungatella}.
Communities with greater abundances of \emph{Hungatella},
\emph{Eggerthella}, \emph{Bifidobacterium}, \emph{Duncaniella} and
\emph{Neisseria} were more likely to have high histopathologic scores.
These models have shown that the relative abundance of bacterial
populations and their relationship to each other could be used to
predict the variation in moribundity, histopathologic score, and toxin
activity of CDI.

\textbf{Disease severity consistent by donor community across isolates.}
Since Carlson et al.~showed \emph{C. difficile} isolates with lower
germination rates were more likely to result in severe disease (27), we
next tested whether the effect of the community on CDI outcomes was
conserved across \emph{C. difficile} isolates with higher rates of
germination. We colonized mice with human fecal slurry of either Donor
N7, N6, or M4 and allowed two weeks for the community to establish. We
then challenged the mice with a different \emph{C. difficile} isolate
and all mice were infected. The moribundity and histopathologic score
outcomes when challenged with a different \emph{C. difficile} human
isolate were comparable to the outcomes of challenges with \emph{C.
difficile} RT027 human isolate 431 by their donor group (Figure 6). For
the isolates and communities we tested, the effect of the gut bacterial
community on disease severity was greater than the effect of the
variation in these isolates.

@@

Challenging mice colonized with different human fecal communities with a
single \emph{C. difficile} isolate, RT027 isolate 431, demonstrated that
variation in the gut bacteria affects the \emph{C. difficile} infection
disease severity. Our analysis revealed an association between the
relative abundance of bacterial community members and disease severity.
We used the relative abundance patterns of community members associated
with disease outcomes to select donor communities to test whether the
disease severity was conserved when challenged with a different \emph{C.
difficile} isolate. The disease outcomes from the challenge with the
second \emph{C. difficile} isolate matched the outcomes of the challenge
with \emph{C. difficile} RT027 isolate 431. Therefore, the effect of the
community variation on disease severity \emph{in vivo} was greater than
the effect of \emph{C. difficile} isolate variation (27, 28). Overall,
these results have demonstrated that the bacterial community contributed
to the disease severity of \emph{C. difficile} infection.
